---
title: "PVN Female Avp Neurons"
author: "Sinead Archdeacon"
date: "2026-04-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This section subsets female Avp+ neurons based on the previous annotations and runs DE analysis to compare gene expression for stress and control samples.

```{r Environment setup}

# Increase capacity of R: necessary at the start of every session
options(future.globals.maxSize = 60000 * 1024^2)

# Load libraries
library(Seurat)
library(SeuratObject)
library(future.apply)
library(harmony)
library(glmGamPoi)
library(sctransform)
library(biomaRt)
library(viridis)
library(tidyverse)
library(kableExtra)
library(edgeR)
library(HGNChelper)
library(ggplot2)
library(DESeq2)
library(Matrix)
library(data.table)
library(S4Vectors)
```

## Read in Avp+ subset

```{r read in data}

AvpSubset.35_0.75 <- readRDS(file = "C:/shares/btrainor/Sinead/GitHub/AvpSubset.35_0.75.rds")
```

## Subset Avp neurons

```{r subset Avp neurons}

# Subset for just neurons based on cell type 1
AvpNeurons <- subset(AvpSubset.35_0.75, Assigned_Cell_Type == "Neuron_Score")

# Re-process the subset
DefaultAssay(AvpNeurons) <- "SCT"
VariableFeatures(AvpNeurons) <- SelectIntegrationFeatures(object.list = list(AvpNeurons), nfeatures = 2000)
ElbowPlot(AvpNeurons, 50)
AvpNeurons <- RunPCA(AvpNeurons, features = VariableFeatures(AvpNeurons), verbose = FALSE)
AvpNeurons <- RunHarmony(AvpNeurons, assay.use = "SCT", group.by.vars = "dataset")
AvpNeurons.40 <- FindNeighbors(AvpNeurons, reduction = "harmony", dims = 1:40)
AvpNeurons.40_0.3 <- FindClusters(AvpNeurons.40, resolution = 0.3)
AvpNeurons.40_0.3 <- RunUMAP(AvpNeurons.40_0.3, reduction = "harmony", dims = 1:40)

# 2011 females, 490 males
```

Update metadata columns to show technical and biological replicates.

```{r update metadata columns}

colnames(AvpNeurons.40_0.3@meta.data)[colnames(AvpNeurons.40_0.3@meta.data) == "assignment"] <- "bio_replicate"
AvpNeurons.40_0.3$tech_replicate <- gsub(".*(\\d)$", "\\1", AvpNeurons.40_0.3$dataset)

AvpNeurons.40_0.3@meta.data$bio_replicate <- as.character(AvpNeurons.40_0.3@meta.data$bio_replicate)
AvpNeurons.40_0.3@meta.data$bio_replicate[AvpNeurons.40_0.3@meta.data$tech_replicate == "2" & AvpNeurons.40_0.3@meta.data$bio_replicate == "0" & AvpNeurons.40_0.3@meta.data$sex == "female"] <- "4"
AvpNeurons.40_0.3@meta.data$bio_replicate[AvpNeurons.40_0.3@meta.data$tech_replicate == "2" & AvpNeurons.40_0.3@meta.data$bio_replicate == "1" & AvpNeurons.40_0.3@meta.data$sex == "female"] <- "5"
AvpNeurons.40_0.3@meta.data$bio_replicate[AvpNeurons.40_0.3@meta.data$tech_replicate == "2" & AvpNeurons.40_0.3@meta.data$bio_replicate == "2" & AvpNeurons.40_0.3@meta.data$sex == "female"] <- "6"
AvpNeurons.40_0.3@meta.data$bio_replicate[AvpNeurons.40_0.3@meta.data$tech_replicate == "2" & AvpNeurons.40_0.3@meta.data$bio_replicate == "3" & AvpNeurons.40_0.3@meta.data$sex == "female"] <- "7"

# Remove rows where bio_replicate is NA
AvpNeurons.40_0.3 <- AvpNeurons.40_0.3[, !is.na(AvpNeurons.40_0.3@meta.data$bio_replicate)]

# Check results
table(is.na(AvpNeurons.40_0.3@meta.data$bio_replicate))
table(AvpNeurons.40_0.3@meta.data$bio_replicate)
table(AvpNeurons.40_0.3@meta.data$sex)

# Save progress
saveRDS(AvpNeurons.40_0.3, file = "C:/shares/btrainor/Sinead/GitHub/AvpNeurons.40_0.3.rds")
```

## Subset female Avp neurons

```{r subset female Avp neurons}

female.avp.neurons <- subset (AvpNeurons.40_0.3, sex == "female")

# Re-process the subset
DefaultAssay(female.avp.neurons) <- "SCT"
VariableFeatures(female.avp.neurons) <- SelectIntegrationFeatures(object.list = list(female.avp.neurons), 
                                                                  nfeatures = 2000)
female.avp.neurons <- RunPCA(female.avp.neurons, features = VariableFeatures(female.avp.neurons), verbose = FALSE)
female.avp.neurons <- RunHarmony(female.avp.neurons, assay.use = "SCT", group.by.vars = "dataset")
ElbowPlot(female.avp.neurons, 50)
female.avp.neurons.30 <- FindNeighbors(female.avp.neurons, reduction = "harmony", dims = 1:30)
female.avp.neurons.30_0.3 <- FindClusters(female.avp.neurons.30, resolution = 0.3)
female.avp.neurons.30_0.3 <- RunUMAP(female.avp.neurons.30_0.3, reduction = "harmony", dims = 1:30)

# check metadata
table(female.avp.neurons.30_0.3@meta.data$bio_replicate)
table(female.avp.neurons.30_0.3@meta.data$tech_replicate)

# join counts layer
DefaultAssay(female.avp.neurons.30_0.3) <- "RNA"
female.avp.neurons.30_0.3 <- JoinLayers(object = female.avp.neurons.30_0.3, assay = "RNA", layers = "counts")

# save progress
saveRDS(female.avp.neurons.30_0.3, file = "C:/shares/btrainor/Sinead/GitHub/female.avp.neurons.30_0.3.rds")
```

## DE for female Avp neurons

First, structure the counts matrix using Pseudobulk.

```{r Pseudobulk for female Avp neurons}

# Update the sample_identifiers column to reflect updated tech_replicates
female.avp.neurons.30_0.3@meta.data$sample_identifiers <- paste(female.avp.neurons.30_0.3@meta.data$condition, female.avp.neurons.30_0.3@meta.data$tech_replicate, sep = "_")
unique(female.avp.neurons.30_0.3@meta.data$sample_identifiers) # check

# create counts matrix
cts <- GetAssayData(female.avp.neurons.30_0.3, assay = "RNA", layer = "counts")
stopifnot(is(cts, "Matrix"))

# ensure sparse matrix format
cts <- as(cts, "dgCMatrix")
stopifnot(is(cts, "dgCMatrix"))

# convert to dataframe
cts_df <- as.data.frame(as.matrix(cts))
stopifnot(is.data.frame(cts_df)) 

# transpose and convert to dataframe
GEM <- as.data.frame(t(cts_df))
stopifnot(is.data.frame(GEM))

# ensure sample identifiers exist in metadata
stopifnot("sample_identifiers" %in% colnames(female.avp.neurons.30_0.3@meta.data))
GEM$sample_identifiers <- female.avp.neurons.30_0.3@meta.data$sample_identifiers

# aggregate counts
aggr <- aggregate(GEM[, -ncol(GEM)], by = list(GEM$sample_identifiers), FUN = sum)
stopifnot(is.data.frame(aggr))

# ensure proper rownames
rownames(aggr) <- aggr$Group.1
aggr$Group.1 <- NULL

# clean up environment
rm(cts, cts_df, GEM)

# ensure all columns are numeric
aggr <- as.data.frame(lapply(aggr, as.numeric), row.names = rownames(aggr))
stopifnot(all(sapply(aggr, is.numeric)))  

# convert "aggr" to "cts_matrix"
cts_matrix <- as.matrix(aggr)

# ensure correct format
stopifnot(is.matrix(cts_matrix)) 
stopifnot(mode(cts_matrix) == "numeric")
cat("\nChecking structure of cts_matrix...\n")
dim(cts_matrix)
rownames(cts_matrix)[1:10]  # should be gene names
colnames(cts_matrix)[1:10]  # should be sample names
cts_matrix <- (t(cts_matrix))
rownames(cts_matrix)[1:10]  # should be gene names
colnames(cts_matrix)[1:10]  # should be sample names
mode(cts_matrix)  # should return "numeric"
class(cts_matrix)  # should return "matrix"

# filter lowly expressed genes
cts_matrix <- cts_matrix[rowSums(cts_matrix) > 5, ]
stopifnot(nrow(cts_matrix) > 0)  # ensure some genes remain
```

Next, create a column data object that is compatible with the counts matrix.

```{r create coldata}

coldata <- data.frame(
  row.names = colnames(cts_matrix),
  condition = gsub("_\\d+", "", colnames(cts_matrix)),
  tech_replicate = gsub("^.*_(\\d+)$", "\\1", colnames(cts_matrix)))

# convert to factors
coldata$condition <- factor(coldata$condition)
coldata$tech_replicate <- factor(coldata$tech_replicate)

# ensure proper structure
stopifnot(is.data.frame(coldata))  # should be a dataframe
stopifnot(all(rownames(coldata) == colnames(cts_matrix)))  # row names must match column names of cts_matrix
```

Finalize the counts matrix and coldata formats.

```{r finalize counts matrix and coldata}

# convert counts matrix to sparse matrix
cts_matrix <- as(cts_matrix, "dgCMatrix")  
stopifnot(is(cts_matrix, "dgCMatrix")) 

stopifnot(ncol(cts_matrix) > 0)  # ensure some samples remain
str(coldata)  # should contain 'condition' and 'bio_replicate'
all(rownames(coldata) == colnames(cts_matrix))  # should return TRUE
levels(coldata$condition)  # should return "control" and "stress"
```

Then create a DESeq2 dataset (dds) using the counts matrix and coldata

```{r create dds}

dds <- tryCatch({
  DESeq2::DESeqDataSetFromMatrix(countData = cts_matrix, colData = coldata, design = tech_replicate ~ condition)
}, error = function(e) {
  cat("Error in DESeqDataSetFromMatrix:\n", conditionMessage(e), "\n")
  stop(e)
})

stopifnot(is(dds, "DESeqDataSet"))
stopifnot("condition" %in% colnames(coldata))  
coldata <- DataFrame(coldata) # ensure coldata is a proper DataFrame
```

Run DE, filter and save results
*Note:* there were no DEs here for padj <0.05

```{r run DE}

dds <- DESeq(dds)
summary(dds)

# inspect results
res <- results(dds)
stopifnot(is(res, "DESeqResults"))
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)
res_df$DE <- ifelse(res_df$log2FoldChange < 0, "down", "up")
summary(res$log2FoldChange)
summary(res$padj)

# save results with all p-values (remove LOC genes)
res_df_noLOC <- res_df %>%
  filter(!grepl("LOC", gene))
all_genes_fem_avp <- res_df_noLOC
write.csv(all_genes_fem_avp, "C:/shares/btrainor/Sinead/GitHub/all_genes_fem_avp.csv") 

# save significant results (remove LOC genes)
results_filtered_pvalue <- res_df %>% 
  filter(pvalue < 0.05, !grepl("LOC", gene))
write.csv(results_filtered_pvalue, "C:/shares/btrainor/Sinead/GitHub/female.avp.neurons.30_0.3_results_pval.csv")
```






















