---
title: "PVN analysis 2024-2025: cell type assignment"
author: "Sinead Archdeacon"
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This section of the workflow annotates the Avp+ and Oxt+ cells with broad cell type classes, so neurons can be identified.

```{r Environment setup}

# Increase capacity of R: necessary at the start of every session
options(future.globals.maxSize = 60000 * 1024^2)

# Load libraries
library(Seurat)
library(SeuratObject)
library(future.apply)
library(harmony)
library(glmGamPoi)
library(sctransform)
library(biomaRt)
library(viridis)
library(tidyverse)
library(kableExtra)
library(edgeR)
library(HGNChelper)
library(ggplot2)
```

## Read in the data

```{r read in data}

AvpSubset.35_0.75 <- readRDS(file = "C:/shares/btrainor/Sinead/GitHub/AvpSubset.35_0.75.rds")
OxtSubset.45_0.4 <- readRDS(file = "C:/shares/btrainor/Sinead/GitHub/AvpSubset.45_0.4.rds")
```

## Define cell type markers

```{r cell type markers}

neuron_genes <- c("Snap25", "Gria1", "Rbfox3")
astrocyte_genes <- c("Gfap", "Aldh1l1", "S100b", "Sox9", "Aqp4", "Slc1a3", "Ndrg2")
oligodendrocyte_genes <- c("Mbp", "Mog", "Plp1", "Mal", "Mobp")
microglial_genes <- c("P2ry12", "Aif1", "Tmem119", "Cx3cr1")
endothelial_genes <- c("Pecam1","Flt1", "Pltp", "Epas1", "Flt1", "Lrp8", "Nos3")
pericyte_genes <- c("Pdgfrb", "Acta2", "Des", "Anpep", "Cspg4")
ependymal_genes <- c("Foxj1", "Ttr", "Vim")

all_genes <- list(
  neuron = neuron_genes,
  astrocyte = astrocyte_genes,
  oligodendrocyte = oligodendrocyte_genes,
  microglia = microglial_genes,
  endothelial = endothelial_genes,
  pericyte = pericyte_genes,
  ependymal = ependymal_genes
)
```

## Assign module scores
Assign module scores for each cell type in both subsets.

```{r module scores}

DefaultAssay(AvpSubset.35_0.75) <- "SCT"
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(neuron_genes), 
                                            name = "Neuron_Score")
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(astrocyte_genes), 
                                            name = "Astrocyte_Score")
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(oligodendrocyte_genes), 
                                            name = "Oligodendrocyte_Score")
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(microglial_genes), 
                                            name = "Microglial_Score")
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(endothelial_genes), 
                                            name = "Endothelial_Score")
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(pericyte_genes), 
                                            name = "Pericyte_Score")
AvpSubset.35_0.75 <- AddModuleScore(AvpSubset.35_0.75, features = list(ependymal_genes), 
                                            name = "Ependymal_Score")

DefaultAssay(OxtSubset.45_0.4) <- "SCT"
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(neuron_genes), 
                                           name = "Neuron_Score")
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(astrocyte_genes), 
                                   name = "Astrocyte_Score")
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(oligodendrocyte_genes), 
                                           name = "Oligodendrocyte_Score")
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(microglial_genes), 
                                   name = "Microglial_Score")
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(endothelial_genes), 
                                   name = "Endothelial_Score")
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(pericyte_genes), 
                                   name = "Pericyte_Score")
OxtSubset.45_0.4 <- AddModuleScore(OxtSubset.45_0.4, features = list(ependymal_genes),
                                   name = "Ependymal_Score")

```

## Cell type 0
Assign "cell type 0" for each subset based on module scores.

```{r assign cell type 0}

AvpSubset.35_0.75@meta.data
meta_data <-  AvpSubset.35_0.75@meta.data
scores_df <- data.frame(
  Neuron_Score = meta_data$Neuron_Score1,
  Astrocyte_Score = meta_data$Astrocyte_Score1,
  Microglial_Score = meta_data$Microglial_Score1,
  Endothelial_Score = meta_data$Endothelial_Score1,
  Oligodendrocyte_Score = meta_data$Oligodendrocyte_Score1,
  Pericyte_Score = meta_data$Pericyte_Score1,
  Ependymal_Score = meta_data$Ependymal_Score1
)

classify_cell_type <- function(row) {
  max_score <- max(row, na.rm = TRUE)
  if (is.na(max_score)) return("Unknown")
  close_scores <- sum(abs(row - max_score) <= 0.025, na.rm = TRUE)
  if (close_scores > 1) return("Other")
  return(names(row)[which.max(row)])
}

AvpSubset.35_0.75$Cell_Type_0 <- apply(scores_df, 1, classify_cell_type)

OxtSubset.45_0.4@meta.data
meta_data <-  OxtSubset.45_0.4@meta.data
scores_df <- data.frame(
  Neuron_Score = meta_data$Neuron_Score1,
  Astrocyte_Score = meta_data$Astrocyte_Score1,
  Microglial_Score = meta_data$Microglial_Score1,
  Endothelial_Score = meta_data$Endothelial_Score1,
  Oligodendrocyte_Score = meta_data$Oligodendrocyte_Score1,
  Pericyte_Score = meta_data$Pericyte_Score1,
  Ependymal_Score = meta_data$Ependymal_Score1
)

classify_cell_type <- function(row) {
  max_score <- max(row, na.rm = TRUE)
  if (is.na(max_score)) return("Unknown")
  close_scores <- sum(abs(row - max_score) <= 0.025, na.rm = TRUE)
  if (close_scores > 1) return("Other")
  return(names(row)[which.max(row)])
}

OxtSubset.45_0.4$Cell_Type_0 <- apply(scores_df, 1, classify_cell_type)
```

Visualize cell type 0

```{r visualize cell types}

DimPlot(AvpSubset.35_0.75, 
         reduction = "umap", 
         group.by = "Cell_Type_0", 
         label = FALSE, 
         pt.size = 1) + 
  guides(color = guide_legend(title = "Cell Type")) + 
  theme(legend.position = "right")
# 1855 Avp neurons

DimPlot(OxtSubset.45_0.4, 
                      reduction = "umap", 
                      group.by = "Cell_Type_0", 
                      label = FALSE, 
                      pt.size = 1) + 
        guides(color = guide_legend(title = "Cell Type")) + 
        theme(legend.position = "right")
# 4684 Oxt neurons
```

## Assign Cell type 1
To get more tidy clustering, I then assigned "cell type 1" by cluster, using whichever cell type 0 assignment was most common in that cluster (essentially, by-cluster voting).

```{r assign Avp cell type 1}

# Step 1: summarize the cell type assignments by cluster

DefaultAssay(AvpSubset.35_0.75) <- "SCT"
cell_type_summary <-  AvpSubset.35_0.75@meta.data %>%
  group_by(seurat_clusters, Cell_Type_0) %>%
  summarise(cell_count = n(), .groups = 'drop') %>%
  group_by(seurat_clusters) %>%
  mutate(
    total_cells = sum(cell_count),
    percent = (cell_count / total_cells) * 100
  ) %>%
  ungroup()

# Step 2: create a lookup vector for cluster-level assignments

cluster_assignments <- cell_type_summary %>%
  group_by(seurat_clusters) %>%
  summarise(
    Assigned_Cell_Type = Cell_Type_0[which.max(cell_count)], # Assign the most frequent cell type
    .groups = 'drop'
  ) %>%
  # Ensure clusters are in order
  arrange(seurat_clusters)

# View the resulting cluster assignments
print(cluster_assignments)

# Step 3: ensure all clusters are included

all_clusters <- unique(AvpSubset.35_0.75@meta.data$seurat_clusters)
missing_clusters <- setdiff(all_clusters, cluster_assignments$seurat_clusters)

if (length(missing_clusters) > 0) {
  # Add missing clusters with "Other" assigned
  missing_assignments <- tibble(
    seurat_clusters = missing_clusters,
    Assigned_Cell_Type = "Other"
  )
  cluster_assignments <- bind_rows(cluster_assignments, missing_assignments)
}

# Step 4: map Assigned_Cell_Type to metadata without using join

# Create a named vector for fast mapping
assignment_vector <- setNames(cluster_assignments$Assigned_Cell_Type, 
                              cluster_assignments$seurat_clusters)

# Step 5: assign cell types directly by indexing

AvpSubset.35_0.75@meta.data$Assigned_Cell_Type <- 
  assignment_vector[as.character(AvpSubset.35_0.75@meta.data$seurat_clusters)]

# Step 6: check if any NA values were introduced (shouldn't happen)

if (any(is.na(AvpSubset.35_0.75@meta.data$Assigned_Cell_Type))) {
  stop("Some clusters were not assigned a cell type.")
}

# Step 7: view counts of cells by assigned cell type

cell_type_counts <- AvpSubset.35_0.75@meta.data %>%
  group_by(Assigned_Cell_Type) %>%
  summarise(Cell_Count = n(), .groups = 'drop')

# Print the cell type counts
print(cell_type_counts) # 2501 Avp neurons
```

```{r assign Oxt cell type 1}

# Step 1: summarize the cell type assignments by cluster

DefaultAssay(OxtSubset.45_0.4) <- "SCT"
cell_type_summary <-  OxtSubset.45_0.4@meta.data %>%
  group_by(seurat_clusters, Cell_Type_0) %>%
  summarise(cell_count = n(), .groups = 'drop') %>%
  group_by(seurat_clusters) %>%
  mutate(
    total_cells = sum(cell_count),
    percent = (cell_count / total_cells) * 100
  ) %>%
  ungroup()

# Step 2: create a lookup vector for cluster-level assignments

cluster_assignments <- cell_type_summary %>%
  group_by(seurat_clusters) %>%
  summarise(
    Assigned_Cell_Type = Cell_Type_0[which.max(cell_count)], # Assign the most frequent cell type
    .groups = 'drop'
  ) %>%
  # Ensure clusters are in order
  arrange(seurat_clusters)

# View the resulting cluster assignments
print(cluster_assignments)

# Step 3: ensure all clusters are included

all_clusters <- unique(OxtSubset.45_0.4@meta.data$seurat_clusters)
missing_clusters <- setdiff(all_clusters, cluster_assignments$seurat_clusters)

if (length(missing_clusters) > 0) {
  # Add missing clusters with "Other" assigned
  missing_assignments <- tibble(
    seurat_clusters = missing_clusters,
    Assigned_Cell_Type = "Other"
  )
  cluster_assignments <- bind_rows(cluster_assignments, missing_assignments)
}

# Step 4: map Assigned_Cell_Type to metadata without using join

# Create a named vector for fast mapping
assignment_vector <- setNames(cluster_assignments$Assigned_Cell_Type, 
                              cluster_assignments$seurat_clusters)

# Step 5: assign cell types directly by indexing

OxtSubset.45_0.4@meta.data$Assigned_Cell_Type <- 
  assignment_vector[as.character(OxtSubset.45_0.4@meta.data$seurat_clusters)]

# Step 6: check if any NA values were introduced (shouldn't happen)

if (any(is.na(OxtSubset.45_0.4@meta.data$Assigned_Cell_Type))) {
  stop("Some clusters were not assigned a cell type.")
}

# Step 7: view counts of cells by assigned cell type

cell_type_counts <- OxtSubset.45_0.4@meta.data %>%
  group_by(Assigned_Cell_Type) %>%
  summarise(Cell_Count = n(), .groups = 'drop')

# Print the cell type counts
print(cell_type_counts) # 6310 Oxt neurons
```

Visualize cell type 1 (improved compared to cell type 0)

```{r visualize cell type 1}

DimPlot(AvpSubset.35_0.75, 
  reduction = "umap", 
  group.by = "Assigned_Cell_Type", 
  label = FALSE, 
  pt.size = 1) + 
  guides(color = guide_legend(title = "Cell Type")) + 
  theme(legend.position = "right")

DimPlot(OxtSubset.45_0.4, 
  reduction = "umap", 
  group.by = "Assigned_Cell_Type", 
  label = FALSE, 
  pt.size = 1) + 
  guides(color = guide_legend(title = "Cell Type")) + 
  theme(legend.position = "right")
```

```{r save progress}

saveRDS(AvpSubset.35_0.75, file = "C:/shares/btrainor/Sinead/GitHub/AvpSubset.35_0.75.rds")
saveRDS(OxtSubset.45_0.4, file = "C:/shares/btrainor/Sinead/GitHub/OxtSubset.45_0.4.rds")
```









