---
title: "PVN analysis 2024-2025: preprocessing"
author: "Sinead Archdeacon"
date: "2026-04-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This pipeline follows a standard Seurat workflow for loading and preprocessing snRNA-seq data. The data is then subsetted for cells of interest for downstream analysis.

```{r Environment setup}

# Increase capacity of R: necessary at the start of every session
options(future.globals.maxSize = 60000 * 1024^2)

# Load libraries
library(Seurat)
library(SeuratObject)
library(future.apply)
library(harmony)
library(glmGamPoi)
library(sctransform)
library(biomaRt)
library(viridis)
library(tidyverse)
library(kableExtra)
library(edgeR)
library(HGNChelper)
library(ggplot2)
```

## Seurat object creation

First, add the .h5 files to the RStudio environment.

```{r Upload files}

PVNData_malec1 <- Read10X_h5(filename = 'C:/shares/btrainor/Sequencingdata/Resequenced_PVN/PVN_MALE_C1/filtered_feature_bc_matrix.h5', use.names = TRUE, unique.features = TRUE)
PVNData_males1 <- Read10X_h5(filename = 'C:/shares/btrainor/Sequencingdata/Resequenced_PVN/PVN_MALE_S1/filtered_feature_bc_matrix.h5', use.names = TRUE, unique.features = TRUE)
PVNData_fems1 <- Read10X_h5(filename = 'C:/shares/btrainor/Sequencingdata/Resequenced_PVN/AG_BT_PVN_S1/filtered_feature_bc_matrix.h5', use.names = TRUE, unique.features = TRUE)
PVNData_femc1 <- Read10X_h5(filename = 'C:/shares/btrainor/Sequencingdata/Resequenced_PVN/AG_BT_PVN_C1/filtered_feature_bc_matrix.h5', use.names = TRUE, unique.features = TRUE)
PVNData_fems2 <- Read10X_h5(filename = 'C:/shares/btrainor/Sequencingdata/Resequenced_PVN/PVN_FEM_S2/filtered_feature_bc_matrix.h5', use.names = TRUE, unique.features = TRUE)
PVNData_femc2 <- Read10X_h5(filename = 'C:/shares/btrainor/Sequencingdata/Resequenced_PVN/PVN_FEM_C2/filtered_feature_bc_matrix.h5', use.names = TRUE, unique.features = TRUE)
```

Convert to Seurat objects and add relevant metadata columns. The Souporcell information is used to demultiplex the pooled samples.

```{r Create Seurat objects}

PVNSeuratObject_malec1 <- CreateSeuratObject(counts = PVNData_malec1)
PVNSeuratObject_males1 <- CreateSeuratObject(counts = PVNData_males1)
PVNSeuratObject_femc1 <- CreateSeuratObject(counts = PVNData_femc1)
PVNSeuratObject_fems1 <- CreateSeuratObject(counts = PVNData_fems1)
PVNSeuratObject_femc2 <- CreateSeuratObject(counts = PVNData_femc2)
PVNSeuratObject_fems2 <- CreateSeuratObject(counts = PVNData_fems2)

# add sex metadata
PVNSeuratObject_malec1$sex <- 'male'
PVNSeuratObject_males1$sex <- 'male'
PVNSeuratObject_femc1$sex <- 'female'
PVNSeuratObject_femc2$sex <- 'female'
PVNSeuratObject_fems1$sex <- 'female'
PVNSeuratObject_fems2$sex <- 'female'

# add condition metadata
PVNSeuratObject_malec1$condition <- 'control'
PVNSeuratObject_males1$condition <- 'stress'
PVNSeuratObject_femc1$condition <- 'control'
PVNSeuratObject_femc2$condition <- 'control'
PVNSeuratObject_fems1$condition <- 'stress'
PVNSeuratObject_fems2$condition <- 'stress'

# add Souporcell metadata (mouse IDs and doublet status)
  add_souporcell_metadata <- function(seurat_object, souporcell_dir) {
    clusters_path <- file.path(souporcell_dir, "clusters.tsv")
  }

  clusters <- read.table(clusters_path, header = TRUE, stringsAsFactors = FALSE)

  # Convert barcode column to character and trim spaces
  clusters$barcode <- as.character(trimws(clusters$barcode)) 

  # Find matching barcodes
  matching_barcodes <- clusters$barcode %in% colnames(seurat_object) 
  matched_count <- sum(matching_barcodes)
  print(paste("Matching barcodes found:", matched_count))
   if (matched_count == 0) {
      stop("Error: No overlapping cells found! Double-check barcode formatting.")
   }

# Subset only matching barcodes
clusters <- clusters[matching_barcodes, c("barcode", "assignment", "status")] 

# Ensure row names match Seurat object
rownames(clusters) <- clusters$barcode
clusters$barcode <- NULL  # Remove barcode column after setting rownames
  
# Add metadata to Seurat object
seurat_object <- Seurat::AddMetaData(seurat_object, metadata = clusters)
  
return(seurat_object)

# Remove doublets using doublet status metadata column
PVNSeuratObject_malec1 <- subset(PVNSeuratObject_malec1, subset = status == "singlet" & 
                                   assignment %in% c("0", "1", "2", "3"))
PVNSeuratObject_males1 <- subset(PVNSeuratObject_males1, subset = status == "singlet" &
                                   assignment %in% c("0", "1", "2", "3"))
PVNSeuratObject_femc1 <- subset(PVNSeuratObject_femc1, subset = status == "singlet" & 
                                  assignment %in% c("0", "1", "2", "3"))
PVNSeuratObject_femc2 <- subset(PVNSeuratObject_femc2, subset = status == "singlet" & 
                                  assignment %in% c("0", "1", "2", "3"))
PVNSeuratObject_fems1 <- subset(PVNSeuratObject_fems1, subset = status == "singlet" & 
                                  assignment %in% c("0", "1", "2", "3"))
PVNSeuratObject_fems2 <- subset(PVNSeuratObject_fems2, subset = status == "singlet" & 
                                  assignment %in% c("0", "1", "2", "3"))
# Check how many doublets were removed

# Add cell IDS and orig idents metadata
PVNSeuratObject_malec1 <- RenameCells(PVNSeuratObject_malec1, add.cell.id = "malec1_")
PVNSeuratObject_males1 <- RenameCells(PVNSeuratObject_males1, add.cell.id = "males1_")
PVNSeuratObject_femc1 <- RenameCells(PVNSeuratObject_femc1, add.cell.id = "femc1_")
PVNSeuratObject_femc2 <- RenameCells(PVNSeuratObject_femc2, add.cell.id = "femc2_")
PVNSeuratObject_fems1 <- RenameCells(PVNSeuratObject_fems1, add.cell.id = "fems1_")
PVNSeuratObject_fems2 <- RenameCells(PVNSeuratObject_fems2, add.cell.id = "fems2_")

PVNSeuratObject_malec1$orig.ident <- "malec1"
PVNSeuratObject_males1$orig.ident <- "males1"
PVNSeuratObject_femc1$orig.ident <- "femc1"
PVNSeuratObject_femc2$orig.ident <- "femc2"
PVNSeuratObject_fems1$orig.ident <- "fems1"
PVNSeuratObject_fems2$orig.ident <- "fems2"

# Add 'dataset' metadata (will help with pseudobulking later on)
PVNSeuratObject_malec1$dataset <- "male_c1"
PVNSeuratObject_males1$dataset <- "male_s1"
PVNSeuratObject_femc1$dataset <- "female_c1"
PVNSeuratObject_femc2$dataset <- "female_c2"
PVNSeuratObject_fems1$dataset <- "female_s1"
PVNSeuratObject_fems2$dataset <- "female_s2"

```

## Quality control/filtering
Refer to 1-preprocessing.Rmd in BNST_analysis_2025 repo for details on method used.

```{r QC/filtering}

VlnPlot(PVNSeuratObject_malec1, features = c("nFeature_RNA", "nCount_RNA"), ncol = 3)
FeatureScatter(PVNSeuratObject_malec1, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = 'lm')
PVNSeuratObject_malec1 <- subset(PVNSeuratObject_malec1, subset = nFeature_RNA > 1000 & nFeature_RNA < 10000 & nCount_RNA < 40000)
```

```{r QC males1}

VlnPlot(PVNSeuratObject_males1, features = c("nFeature_RNA", "nCount_RNA" ), ncol = 3)
FeatureScatter(PVNSeuratObject_males1, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = 'lm')
PVNSeuratObject_males1 <- subset(PVNSeuratObject_males1, subset = nFeature_RNA > 1000 & nFeature_RNA < 10000 & nCount_RNA < 40000)
```

```{r QC femc1}

VlnPlot(PVNSeuratObject_femc1, features = c("nFeature_RNA", "nCount_RNA" ), ncol = 3)
FeatureScatter(PVNSeuratObject_femc1, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = 'lm')
PVNSeuratObject_femc1 <- subset(PVNSeuratObject_femc1, subset = nFeature_RNA > 1000 & nFeature_RNA < 10000 & nCount_RNA < 30000)
```

```{r QC femc2}

VlnPlot(PVNSeuratObject_femc2, features = c("nFeature_RNA", "nCount_RNA" ), ncol = 3)
FeatureScatter(PVNSeuratObject_femc2, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = 'lm')
PVNSeuratObject_femc2 <- subset(PVNSeuratObject_femc2, subset = nFeature_RNA > 1000 & nFeature_RNA < 10000 & nCount_RNA < 40000)
```

```{r QC fems1}

VlnPlot(PVNSeuratObject_fems1, features = c("nFeature_RNA", "nCount_RNA" ), ncol = 3)
FeatureScatter(PVNSeuratObject_fems1, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = 'lm')
PVNSeuratObject_fems1 <- subset(PVNSeuratObject_fems1, subset = nFeature_RNA > 1000 & nFeature_RNA < 10000 & nCount_RNA < 35000)
```

```{r QC fems2}

VlnPlot(PVNSeuratObject_fems2, features = c("nFeature_RNA", "nCount_RNA" ), ncol = 3)
FeatureScatter(PVNSeuratObject_fems2, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + geom_smooth(method = 'lm')
PVNSeuratObject_fems2 <- subset(PVNSeuratObject_fems2, subset = nFeature_RNA > 1000 & nFeature_RNA < 10000 & nCount_RNA < 40000)
```

## Normalize RNA assays for each object

```{r Normalize RNA assays}

PVNSeuratObject_malec1 <- NormalizeData(PVNSeuratObject_malec1, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
PVNSeuratObject_malec1 <- FindVariableFeatures(PVNSeuratObject_malec1, selection.method = "vst", nfeatures = 2000, assay = "RNA")
all.genes <- rownames(PVNSeuratObject_malec1)
PVNSeuratObject_malec1 <- ScaleData(PVNSeuratObject_malec1, features = all.genes, assay = "RNA")

PVNSeuratObject_males1 <- NormalizeData(PVNSeuratObject_males1, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
PVNSeuratObject_males1 <- FindVariableFeatures(PVNSeuratObject_males1, selection.method = "vst", nfeatures = 2000, assay = "RNA")
all.genes <- rownames(PVNSeuratObject_males1)
PVNSeuratObject_males1 <- ScaleData(PVNSeuratObject_males1, features = all.genes, assay = "RNA")

PVNSeuratObject_femc1 <- NormalizeData(PVNSeuratObject_femc1, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
PVNSeuratObject_femc1 <- FindVariableFeatures(PVNSeuratObject_femc1, selection.method = "vst", nfeatures = 2000, assay = "RNA")
all.genes <- rownames(PVNSeuratObject_femc1)
PVNSeuratObject_femc1 <- ScaleData(PVNSeuratObject_femc1, features = all.genes, assay = "RNA")

PVNSeuratObject_femc2 <- NormalizeData(PVNSeuratObject_femc2, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
PVNSeuratObject_femc2 <- FindVariableFeatures(PVNSeuratObject_femc2, selection.method = "vst", nfeatures = 2000, assay = "RNA")
all.genes <- rownames(PVNSeuratObject_femc2)
PVNSeuratObject_femc2 <- ScaleData(PVNSeuratObject_femc2, features = all.genes, assay = "RNA")

PVNSeuratObject_fems1 <- NormalizeData(PVNSeuratObject_fems1, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
PVNSeuratObject_fems1 <- FindVariableFeatures(PVNSeuratObject_fems1, selection.method = "vst", nfeatures = 2000, assay = "RNA")
all.genes <- rownames(PVNSeuratObject_fems1)
PVNSeuratObject_fems1 <- ScaleData(PVNSeuratObject_fems1, features = all.genes, assay = "RNA")

PVNSeuratObject_fems2 <- NormalizeData(PVNSeuratObject_fems2, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
PVNSeuratObject_fems2 <- FindVariableFeatures(PVNSeuratObject_fems2, selection.method = "vst", nfeatures = 2000, assay = "RNA")
all.genes <- rownames(PVNSeuratObject_fems2)
PVNSeuratObject_fems2 <- ScaleData(PVNSeuratObject_fems2, features = all.genes, assay = "RNA")
```

## Combine all Seurat objects and run SCTransform

```{r SCTransform}

# Create a list of Seurat objects
PVN.list <- list(
  malec1 = PVNSeuratObject_malec1,
  males1 = PVNSeuratObject_males1,
  femc1 = PVNSeuratObject_femc1,
  femc2 = PVNSeuratObject_femc2,
  fems1 = PVNSeuratObject_fems1,
  fems2 = PVNSeuratObject_fems2
  )

# Set a larger limit (4 GB) before running SCT
options(future.globals.maxSize = 4e9) 

# Clean up environment
rm(PVNData_femc1, PVNData_femc2, PVNData_fems1, PVNData_fems2, PVNData_malec1, PVNData_males1, PVNSeuratObject_femc1, PVNSeuratObject_femc2, PVNSeuratObject_fems1, PVNSeuratObject_fems2, PVNSeuratObject_malec1, PVNSeuratObject_males1)

plan("multicore", workers = 6) 

# Run SCT (this step takes a long time)
PVN.list <- future_lapply(PVN.list, function(x) {
  Seurat::SCTransform(x, method = "glmGamPoi", 
                      return.only.var.genes = FALSE,  # Keep all genes to avoid missing columns
                      variable.features.n = 2000,    
                      residual_type = "pearson")  # Ensure residuals are calculated properly
}, future.seed = TRUE)
```

## Integrate all individual Seurat objects

```{r integrate}

var.features <- SelectIntegrationFeatures(object.list = PVN.list, nfeatures = 2000)
PVN.sct <- merge(x = PVN.list[[1]], y = PVN.list[2:length(PVN.list)], merge.data = TRUE)
table(PVN.sct@meta.data$dataset)
PVN.sct@meta.data$dataset <- as.factor(PVN.sct@meta.data$dataset)
DefaultAssay(PVN.sct) <- "SCT"  # Make sure SCT is active
VariableFeatures(PVN.sct) <- SelectIntegrationFeatures(object.list = PVN.list, nfeatures = 2000)
length(VariableFeatures(PVN.sct))
VariableFeatures(PVN.sct) <- unique(c(VariableFeatures(PVN.sct), "Avp", "Oxt"))
length(VariableFeatures(PVN.sct)) #2000
```

## PCA and clustering

```{r PCA and clustering}

PVN.sct <- RunPCA(PVN.sct, features = VariableFeatures(PVN.sct), verbose = FALSE)
PVN.sct <- RunHarmony(PVN.sct, assay.use = "SCT", group.by.vars = "dataset")
ElbowPlot(PVN.sct, ndims= 50)
PVN.sct.35 <- FindNeighbors(PVN.sct, reduction = "harmony", dims = 1:35)

plan("sequential")
PVN.sct.35_0.2 <- FindClusters(PVN.sct.35, resolution = 0.2)  
PVN.sct.35_0.2 <- RunUMAP(PVN.sct.35_0.2, reduction = "harmony", dims = 1:35)
DimPlot(PVN.sct.35_0.2, group.by = c("dataset", "assignment", "sex"), ncol = 3)
DimPlot(PVN.sct.35_0.2, reduction = "umap", group.by = "seurat_clusters")
```

## Subset Avp+ cells

Determine vasopressin expression threshold.

```{r Avp threshold}

# Fetch Avp and Oxt expression values
avp_expression <- FetchData(PVN.sct.35_0.2, vars = "Avp", assay = "SCT")
oxt_expression <- FetchData(PVN.sct.35_0.2, vars = "Oxt", assay = "SCT")

# visualize Avp expression
ggplot(avp_expression, aes_string(x = "Avp")) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") + 
  theme_minimal() +
  labs(title = paste("Avp Expression"),
       x = "Expression Level",
       y = "Number of Cells")

# Set dynamic threshold
avp_threshold <- quantile(avp_expression$Avp, 0.90)  # 1.098612

# Subset Avp+ cells
avp_cells <- WhichCells(PVN.sct.35_0.2, expression = Avp > avp_threshold & Avp > Oxt)
AvpSubset <- subset(PVN.sct, cells = avp_cells) # 4897 cells
```

Re-normalize and subcluster the Avp subset.

```{r re-normalize Avp subset}

DefaultAssay(AvpSubset) <- "SCT"
VariableFeatures(AvpSubset) <- SelectIntegrationFeatures(object.list = list(AvpSubset), nfeatures = 2000)

AvpSubset <- RunPCA(AvpSubset, features = VariableFeatures(AvpSubset), verbose = FALSE)
AvpSubset <- RunHarmony(AvpSubset, assay.use = "SCT", group.by.vars = "dataset")
AvpSubset.35 <- FindNeighbors(AvpSubset, reduction = "harmony", dims = 1:35)
AvpSubset.35_0.75 <- FindClusters(AvpSubset.35, resolution = 0.75)
AvpSubset.35_0.75 <- RunUMAP(AvpSubset, reduction = "harmony", dims = 1:35)
DimPlot(AvpSubset.35_0.75, reduction = "umap", group.by = "seurat_clusters")
```

Further visualize Avp expression in the Avp subset.

```{r visualization}

DefaultAssay(AvpSubset.35_0.75) <- "RNA"

FeaturePlot(
  AvpSubset.35_0.75,
  features = "Avp",
  reduction = "umap",
  pt.size = 1    # Adjust point size as needed
  ) +
  ggtitle("UMAP by Avp Expression")

VlnPlot(
  AvpSubset.35_0.75,
  features = "Avp", assay = "RNA",
  group.by = "seurat_clusters",  # Adjust if using a different cluster column
  pt.size = 0  # Set to a small value if you want to show individual points
  ) +
  ggtitle("Avp Expression by Cluster") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Save Avp cells subset

```{r save progress}

saveRDS(AvpSubset.35_0.75, file = "C:/shares/btrainor/Sinead/GitHub/AvpSubset.35_0.75.rds")
```

## Subset Oxt+ cells
Determine oxytocin expression threshold.

```{r oxytocin threshold}

# visualize Oxt expression
ggplot(oxt_expression, aes_string(x = "Oxt")) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") + 
  theme_minimal() +
  labs(title = paste("Oxt Expression"),
       x = "Expression Level",
       y = "Number of Cells")

# Subset Oxt+ cells
oxt_cells <- WhichCells(PVN.sct.35_0.2, expression = Oxt > 0.5 & Oxt > Avp)
OxtSubset <- subset(PVN.sct, cells = oxt_cells) # 9371 cells
```

Re-normalize and subcluster the Oxt subset.

```{r re-normalize Oxt subset}

DefaultAssay(OxtSubset) <- "SCT"
VariableFeatures(OxtSubset) <- SelectIntegrationFeatures(object.list = list(OxtSubset), nfeatures = 2000)

OxtSubset <- RunPCA(OxtSubset, features = VariableFeatures(OxtSubset), verbose = FALSE)
OxtSubset <- RunHarmony(OxtSubset, assay.use = "SCT", group.by.vars = "dataset")
ElbowPlot(OxtSubset, ndims= 50)
OxtSubset.45 <- FindNeighbors(OxtSubset, reduction = "harmony", dims = 1:45)
OxtSubset.45_0.4 <- FindClusters(OxtSubset.45, resolution = 0.4)
OxtSubset.45_0.4 <- RunUMAP(OxtSubset.45_0.4, reduction = "harmony", dims = 1:45)
DimPlot(OxtSubset.45_0.4, reduction = "umap", group.by = "seurat_clusters")
```

Further visualize Oxt expression in the Oxt subset.

```{r visualization}

DefaultAssay(OxtSubset.45_0.4) <- "RNA"
FeaturePlot(
  OxtSubset.45_0.4,
  features = "Oxt",
  reduction = "umap",
  pt.size = 1    
  ) +
  ggtitle("UMAP by Oxt Expression")

VlnPlot(
  OxtSubset.45_0.4,
  features = "Oxt", assay = "RNA",
  group.by = "seurat_clusters",  
  pt.size = 0  
  ) +
  ggtitle("Oxt Expression by Cluster") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r save progress}

saveRDS(OxtSubset.45_0.4, file = "C:/shares/btrainor/Sinead/GitHub/OxtSubset.45_0.4.rds")
```







